---
title: "Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 4"
author: "Matt Shaffer, Ted Pham, Phat Doan"
date: "August 10, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this lab, you are asked to answer the question "Do changes in traffic laws affect traffic fatalities?" To do so, you will conduct the tasks specified below using the data set driving.Rdata, which includes 25 years of data that cover changes in various state drunk driving, seat belt, and speed limit laws.

Specifically, this data set contains data for the 48 continental U.S. states from 1980 through 2004. Various driving laws are indicated in the data set, such as the alcohol level at which drivers are considered legally intoxicated. There are also indicators for "per se" laws-where licenses can be revoked without a trial-and seat belt laws. A few economics and demographic variables are also included. The description of the each ofthe variables in the dataset is come with the dataste.

```{r, warning = F, error = F, message = F}
#rm(list = ls())
library(lme4)
library(stargazer)
library(lattice)
library(arm)
library(Hmisc)
library(psych)
library(car)
library(foreign)
library(gplots)
line_color1 <- rgb(165/255, 203/255, 223/255) 
line_color2 <- rgb(199/255, 37/255, 4/255, 0.5) 
line_color3 <- rgb(200/255, 200/255, 200/255, 0.5) 
line_color4 <- rgb(133/255, 161/255, 198/255, 0.25)
```

```{r}
norm_hist <- function(data.ts, data2.ts=NULL, br=30) {
  if (!is.null(data2.ts)){
    par(mfrow=c(1, 2))
  }
  h <- hist(data.ts, col=rgb(6/255, 57/255, 126/255, 0.4), lwd = 2, breaks = br)
  xfit <- seq(min(data.ts), max(data.ts), length = 40) 
  yfit <- dnorm(xfit, mean = mean(data.ts), sd = sd(data.ts)) 
  yfit <- yfit * diff(h$mids[1:2]) * length(data.ts) 
  lines(xfit, yfit, col = rgb(199/255, 37/255, 4/255, 0.5), lwd = 2)
  
  if (!is.null(data2.ts)){
    h <- hist(data2.ts, col=rgb(6/255, 57/255, 126/255, 0.4), breaks=br)
    xfit <- seq(min(data2.ts), max(data2.ts), length = 40)
    yfit <- dnorm(xfit, mean = mean(data2.ts), sd = sd(data2.ts))
    yfit <- yfit * diff(h$mids[1:2]) * length(data2.ts)

    lines(xfit, yfit, col = rgb(199/255, 37/255, 4/255, 0.5), lwd = 2)
  }
}
```


```{r}
load("driving.Rdata")
drive <- data
```

# Dataset Overview

## Question 1: EDA
Load the data. Provide a description of the basic structure of the dataset, as we have done in throughout the semester. Conduct a very thorough EDA, which should include both graphical and tabular techniques, on the dataset, including both the dependent variable totfatrte and the potential explanatory variables.

### Overview of the dataset
```{r}
#View(drive)
str(drive)
#summary(drive)
```

The dataset has 1200 rows and 56 variables. It is in a panel format at year (1980 through 2004) and state (48 continential states) level. Other variables are:

- sl55: 1 if speed limit == 55

- sl65: 1 if speed limit == 65

- sl70: 1 if speed limit == 70

- sl75: 1 if speed limit == 75

- slnone: 1 if state has no speed limit

- seatbelt: =0 if none, =1 if primary, =2 if secondary

- minage: minimum drinking age

- zerotol: zero tolerance law

- gdl: graduated drivers license law

- bac10: blood alcohol limit .10

- bac08: blood alcohol limit .08

- perse: administrative license revocation (per se law)

- totfat: total traffic fatalities

- nghtfat: total nighttime fatalities

- wkndfat: total weekend fatalities

- totfatpvm: total fatalities per 100 million miles

- nghtfatpvm: nighttime fatalities per 100 million miles

- wkndfatpvm: weekend fatalities per 100 million miles

- statepop: state population

- totfatrte: total fatalities per 100,000 population

- nghtfatrte: nighttime fatalities per 100,000 population

- wkndfatrte: weekend accidents per 100,000 population

- vehicmiles: vehicle miles traveled, billions

- unem: unemployment rate, percent

- perc14_24: percent population aged 14 through 24

- sl70plus: sl70 + sl75 + slnone

- sbprim: 1 if primary seatbelt law

- sbsecon: 1 if secondary seatbelt law

- [d80:d04] =1 if year == [1980:2004]

- vehicmilesp: number of miles driven per capita


```{r}
head(drive, 10)
```


```{r}
tail(drive, 10)
```

```{r}
unique(drive$state)
```
** Check for missing values  **

There are no missing values observed from the data set.
```{r}
sapply(drive, function(x) sum(is.na(x)))
```


### Dependent Variable: totfatrte (total fatalities per 100,000 population)

```{r}
summary(drive$totfatrte)
```

Graph: Bars at top indicates corresponding graph (i.e. countries) from left to right starting on the bottom row 
  -  While there small upticks and variances in the time series, overall we observe the total fatalities per 100,000 population decreases over time from 1980 to 2004
  - 

```{r}
coplot(totfatrte ~ year|state, type="l", data=drive) # Lines
#coplot(totfatrte ~ year|state, type="b", data=drive) # Points and lines
```

```{r}
suppressWarnings(scatterplot(totfatrte ~ year|state, boxplots=FALSE, smooth=TRUE,
                             reg.line=FALSE, data=drive, legend.columns=12,
                             xlab="Year", ylab="Total Fatalities per 100k Population",
                             main="Total Fatalities per 100k Pop by State (1980 - 2004"))
```

```{r}
suppressWarnings(plotmeans(totfatrte ~ state, main="Heterogeineity across States", data=drive,
                           xlab="State", ylab="Total Fatalities per 100k Population"))
```

```{r}
suppressWarnings(plotmeans(totfatrte ~ year, main="Heterogeineity across Years", data=drive,
                           xlab="Year", ylab="Total Fatalities per 100k Population"))
```


```{r}
norm_hist(drive$totfatrte, drive$totfat)
```


### Potential Predictors:

####

```{r}
boxplot(drive$totfatrte ~ drive$year, col=c(line_color1, line_color2))
```


```{r}
xyplot_groups <- function(start_idx, end_idx) {
  xyplot(totfatrte ~ year | state, 
       data = drive[which(drive$state >= start_idx & drive$state <= end_idx), ], 
       aspect = "xy", type = c("g", "p", "r"),
       index.cond = function(x, y) coef(lm(y ~ x))[2], 
       alpha=c(1, 0.5, 0.5), 
       title=drive$state, as.table = TRUE,
       ylab = "total fatalities per 100,000 population",
       xlab = "Year",
       par.settings=simpleTheme(col=line_color1, col.line="red")
       )
}
```

### Bivariate Analysis:

#### Perse


```{r}
xyplot_groups(1, 18)
xyplot_groups(19, 36)
xyplot_groups(37, 52)
```

# Modeling

## Question 2: Simple regression model of *totfatrte* 
How is the our dependent variable of interest totfatrte defined? What is the average of this variable in each of the years in the time period covered in this dataset? Estimate a very simple regression model of totfatrte on dummy variables for the years 1981 through 2004. What does this model explain? Describe what you find in this model. Did driving become safer over this period? Please provide a detailed explanation.

** Global Mean **
```{r}
mean(drive$totfatrte)
```


** Average of `totfatrte` for years `1980`-`2004` **
```{r}
annualMeans <- aggregate(drive$totfatrte, by = list(drive$year), mean)
annualMeans$deviation_from_mean <- annualMeans$x - mean(drive$totfatrte)
annualMeans
```

```{r}
lm.1 <- lm(totfatrte ~ year, data = drive)
lm.2 <- lm(totfatrte ~ as.factor(year), data = drive)
stargazer(lm.1, lm.2, type = "text", summary = FALSE)
```
```{r}

plot(c(1980:2004),lm.2$coefficients)
quick_fit <- lm(lm.2$coefficients~c(1980:2004))
lines(c(1980:2004),fitted(quick_fit))
summary(quick_fit)

```



```{r}
print('bac08')
unique(drive$bac08)
print('bac10')
unique(drive$bac10)
print('perse')
unique(drive$perse)
print('sbprim')
unique(drive$sbprim)
print('sbsecon')
unique(drive$sbsecon)
print('s170plus')
unique(drive$sl70plus)
print('gdl')
unique(drive$gdl)
unique(drive$perc14_24)
hist(drive$vehicmilespc)
```

```{r}
hist(drive$perc14_24)
```


```{r}
hist(drive$gdl)
```


## Question 3:
Expand your model in Exercise 2 by adding variables bac08, bac10, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc, and perhaps transformations of some or all of these variables. Please explain carefully your rationale, which should be based on your EDA, behind any transformation you made. If no transformation is made, explain why transformation is not needed. How are the variables bac8 and bac10 defined? Interpret the coefficients on bac8 and bac10. Do per se laws have a negative effect on the fatality rate? What about having a primary seat belt law? (Note that if a law was enacted sometime within a year the fraction of the year is recorded in place of the zero-one indicator.)

```{r}

model.plus <- lm(totfatrte  ~ bac08 + bac10 + perse + sbprim + sbsecon + sl70plus + gdl + perc14_24 + vehicmilespc, data = drive )

summary(model.plus)

par(mfrow = c(2, 2))
plot(model.plus)

```


```{r}

model.plus <- lm(totfatrte  ~ bac08 + bac10 + perse + sbprim + sbsecon + sl70plus + gdl + perc14_24 + vehicmilespc, data = drive )

summary(model.plus)

```


```{r}
drive$bac08.binary = ifelse(drive$bac08 > 0, 1, 0)
drive$bac10.binary = ifelse(drive$bac10 >0, 1, 0)
drive$perse.binary = ifelse(drive$perse >0, 1, 0)
drive$sl70plus.binary = ifelse(drive$sl70plus > 0, 1, 0)
drive$gdl.binary = ifelse(drive$gdl >0, 1, 0)
```


```{r}

model.plus.binary <- lm(totfatrte  ~ bac08.binary + bac10.binary + perse.binary + sbprim + sbsecon + sl70plus.binary + gdl.binary + perc14_24 + vehicmilespc, data = drive )

summary(model.plus.binary)

par(mfrow = c(2, 2))
plot(model.plus.binary)

```


```{r}
stargazer(model.plus, model.plus.binary, type = "text", summary = FALSE)
```
## Question 4:
Reestimate the model from Exercise 3 using a fixed effects (at the state level) model. How do the coefficients on bac08, bac10, perse, and sbprim compare with the pooled OLS estimates? Which set of estimates do you think is more reliable? What assumptions are needed in each of these models? Are these assumptions reasonable in the current context?


```{r}
library(plm)
drive.panel = plm.data(drive, index = c('state','year'))
model.fe <- plm(totfatrte ~ d80 + d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 + d91 + d92 + d93 + d94 + d00 + d01 + d02 + d03 + d04 +
                          bac08.binary + bac10.binary + perse.binary + sbprim + sbsecon + sl70plus.binary + gdl.binary + perc14_24 + vehicmilespc,
                          data = drive.panel, model = 'within')
summary(model.fe)

```

```{r}

model.fe.nd <- plm(totfatrte ~ 
                          bac08.binary + bac10.binary + perse.binary + sbprim + sbsecon + sl70plus.binary + gdl.binary + perc14_24 + vehicmilespc,
                          data = drive.panel, model = 'within')
summary(model.fe.nd)
```
```{r}

model.rm <- plm(totfatrte ~ d80 + d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 + d91 + d92 + d93 + d94 + d00 + d01 + d02 + d03 + d04 +
                          bac08.binary + bac10.binary + perse.binary + sbprim + sbsecon + sl70plus.binary + gdl.binary + perc14_24 + vehicmilespc,
                          data = drive.panel, model = 'random')
summary(model.rm)
```


```{r}
stargazer(model.fe, model.rm, type='text', summary=FALSE)
```
## Question 5:
Would you perfer to use a random effects model instead of the fixed effects model you build in Exercise 4 ? Why? Why not?

## Question 6:
Suppose that vehicmilespc, the number of miles driven per capita, increases by 1,000. Using the FE estimates, what is the estimated effect on totfatrte? Be sure to interpret the estimate as if explaining to a layperson.

## Question 7:
If there is serial correlation or heteroskedasticity in the idiosyncratic errors of the model, what would be the consequences on the coefficient estimates and their standard errors?


## References:
### Regression models:
Pooled OLS o It considers data as one data set. o No segregation 
Fixed effect model o When we want different intercepts o Change in cross sectional units o No sampling is done 
Random effect Model o When we want to see changes in time and units

### Hypothesis Testing
o Chow Test Ho: Pool is better than fixed Hi: fixed is better than pool 
o Breusch-Pagan test Pool is better than Random Effect Random is better than Pooled 
o Hausman test Random is better than fixed Fixed is better than random