---
title: 'W271 Live Session 11: Analysis of Panel Data 1'
author: "Devesh Tiwari"
date: "7/25/2017"
output:
  pdf_document: default
  html_notebook: default
---

# Main topics covered in Week 12 (Async Unit 11)
    - Introduction to panel data
    - Using OLS regression model on panel data
    - Exploratory panel data analysis
    - Unobserved effect models
    - Pooled OLS models
    - First-Difference models
    - Distributed Lag models

# Readings
**W2016:** Jeffrey Wooldridge. *Introductory Econometrics: A Modern Approach.* 6th edition. Cengage Learning 

    - Ch. 13 (skip 13.4)
    - [package plm](https://cran.r-project.org/web/packages/plm/plm.pdf)
    - [plm vignettes](https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf)

# Agenda for this week's live session:

  1. Questions about ARIMA and SARIMA models
  
  2. Overview of the next unit: Panel data methods
  
  3. EDA of panel and grouped data
  
   Individual-level regression, pooled crossing regression, and first-difference regression

Some start-up codes:
```{r}
#sessionInfo()

# Insert the function to *tidy up* the code when they are printed out
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

# Set Numeric Value Display
# See reference from https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html
options(digits=2) # Set the printed number of digits to 2. Note: It is a suggestion only. Default is 7. 
#options("scipen" = 10)

# Set memory limit
memory.limit(50000000)

# Clean up the workspace before we begin
rm(list = ls())

# Set working directory

# Load libraries
library(car)
library(dplyr)
library(Hmisc)
library(ggplot2)
library(lattice)
library(plm)
```


# 1. Exploratory data analysis of panel data

  - *Estmated Time: Breakout session 25 minutes*
  - *Estmated Time: Class discussion 20 minutes*

In this example, we use five waves of data from "National Youth Survey". Each year, participants in the age of 11, 12, 13, 14, 15 filled out a nine-item instrument designed to assess their tolerance of deviant behiavor such as cheat on tests, sell hard drugs, etc. Response to each item is provided in a 4-point scale (1 = very wrong, 2 = wrong, 3 = a little bit wrong, and 4 = not wrong at all). At each occasion (i.e. wave), the outcome, *TOL*, is computed as the respondent's average across the nine responses. Two potential explanatory variables in this dataset include *male*, which is equal to $1$ if the respondent is *male*, and *exposure*, which is a respondent's estimated proportion of their close friends who were involved in each of the nine activities, measuring on a 5-point scale ($0$ = none, $5$ = all).

**Task 1:** 
    - Import the data as data.frame. 
    - Examine the basic structure of each of the datasets. 
    - Print the person-level dataset. Discuss it.
    - Also, answer "how many male and female in the dataset?" 
    - Construct the correlation matrix of the variables tol11 - tol15. Discuss what you observe in this matrix. Do the corrleation values in the matrix make sense? Why? Why not?
    - Print the person-period level dataset. Discuss it. 
    - How is the person-period level dataset different from that of the person-level dataset?
    - Does the person-level dataset contain exactly the same information as that in the person-period-level dataset as far as *tolerance*, *male*, and *exposure* variables are concern? Why? Why not?
    - Conduct a throughout EDA using ther person-period level dataset
    
```{r}
# YOUR CODE HERE
df <- read.table("tol_person.txt", sep=",", header=T)

df2 <- read.table("tol_person_period.txt", sep=",", header=T)

print(df)
print(df2)

df['male']==0)
```

# 2. Growth-curve Analysis (something not covered in the book and the async)
  - *Estmated Time: Instructor's teaching 10 minutes*

```{r, warning=FALSE}
xyplot(tolerance ~ age | id, data=df2, as.table=T)

# Smooth nonparametric trajectories superimposed on empirical growth plots.
xyplot(tolerance~age | id, data=df2,
      prepanel = function(x, y) prepanel.loess(x, y, family="gaussian"),
      xlab = "Age", ylab = "Tolerance",
      panel = function(x, y) {
      panel.xyplot(x, y)
      panel.loess(x,y, family="gaussian") },
      ylim=c(0, 4), as.table=T)

#plot of the raw data
interaction.plot(df2$age, df2$id, df2$tolerance) 
```


# 3. Build regression models to answer "Does being exposed to friends having deviant behavior increase one's tolerance of deviant behavior?"

  - *Estmated Time: Breakout session 20 minutes*
  - *Estmated Time: Class discussion 15 minutes*

  Task 1: Estimate individual-level regression models. Remember that potential explanatory variables include *time*, *male*, and *exposure*. Can you use all of these variables in the regression models?  Why? Why not? Does this regression help us answer the question? Why? Why not?
  
  Task 2: Answer the question using a pooled-OLS appoach. Interpret the model results.
  
  Task 3: Answer the question using a first-difference approach. Interpret the model results. Conduct EDA on both your dependent variable and explanatory variables first.

```{r}
# YOUR CODE HERE
```

